# Assignment 4: Clustering and classification

In this assignment we will use clustering and classification to analyze a data set included in the MASS package. First, let's take a quick look at what the "Housing Values in Suburbs of Boston" data set contains. We will refer to the data simply as Boston from here on out.

```{r, results=FALSE, message=FALSE, warning=FALSE}
# load in packages for future use, suppress output to keep course diary page tidy
# include MASS package for the Boston data set
library(MASS) # install.packages("MASS")
library(tidyr) # install.packages("tidyverse")
library(corrplot) # install.packages("corrplot")
library(ggplot2) # install.packages("ggplot2")
library(GGally) # install.packages("GGally")
```

```{r}
# load the data
data("Boston")

# explore the dataset
str(Boston)
dim(Boston)
```

We find that the data has 506 observations of 14 variables, 12 of them continuous numerical, and 2 categorical. The column names are not self-explanatory, but more information can be found in: <https://stat.ethz.ch/R-manual/R-devel/library/MASS/html/Boston.html>. Without giving a full list, we can find that the variables include statistics of zoning, housing, and demographics by town area, and importantly measures of air pollution. In the original paper *Harrison, D. and Rubinfeld, D.L. (1978) Hedonic prices and the demand for clean air. J. Environ. Economics and Management 5, 81–102.* the data are used to estimate willingness to "pay for clean air".

## Graphical overview and summaries

To explore the data further, we will present it graphically, and print variable summaries.

```{r}
# variable summaries
summary(Boston)

# scatterplot matrix
pairs(Boston)

# calculate the correlation matrix
cor_matrix <- cor(Boston)

 # visualize the correlation matrix
corrplot(cor_matrix, method="circle", type = "upper", cl.pos = "b", tl.pos = "d", tl.cex = 0.6)
corrplot(cor_matrix, add = TRUE, type = 'lower', method = 'number', col = 'black', diag = FALSE, tl.pos = 'n', cl.pos = 'n', number.cex = 0.6)
```

Just from looking at quartiles in the variable summaries, we can see that many of them are highly skewed. They are also on highly different scales. This could affect data clustering and classification. From the scatterplot matrix, even though it may be difficult to see due to large number of scatterplots, there seem to be some clear divides in data with certain variables. This does not only include the categorical variables of "tract bounds Charles River" (as a dummy variable, the graphical divide does not mean much), and "index of accessibility to radial highways", but also for example the "full-value property-tax rate per $10,000" variable. The correlation plot provides a better view of the correlations between variables. We can see some very strong correlations, notably between indus and nox, indus and dis, indus and tax, nox and age, nox and dis, rm and medv, age and dis, rad and tax, lstat and medv. Some of them are fairly obvious, like correlation between industry and nitrogen oxide concentration, and accessibility to highways and property-tax rate. Also of note is that the "tract bounds Charles River" variable does not correlate much with any other variable.

## Standardization

For classification analyses we will next standardize all variables, and create a categorical variable of the crime variable (for said classification). To run the analyses, the data will be randomly split into a training set with 80% of observations, and a test set with 20% of observations.

```{r}
# standardizing the data set and changing to data frame
boston_scaled <- as.data.frame(scale(Boston))

# summary of scaled data
summary(boston_scaled)

# create a categorical variable crime with crim quantiles as cut points
crime <- cut(boston_scaled$crim, breaks = quantile(boston_scaled$crim), include.lowest = TRUE, label = c("low", "med_low", "med_high", "high"))

# remove original crim from the dataset
boston_scaled <- dplyr::select(boston_scaled, -crim)

# add the new categorical value to scaled data
boston_scaled <- data.frame(boston_scaled, crime)

# number of rows in the Boston dataset 
n <- nrow(boston_scaled)

# choose randomly 80% of the rows
ind <- sample(n,  size = n * 0.8)

# create train set
train <- boston_scaled[ind,]

# create test set 
test <- boston_scaled[-ind,]
```

After scaling, the variables are now centered at 0 and on an equivalent scale.

## Fitting LDA on the train set

Next, we will fit a linear discriminant analysis on the train set with crime rate as the target, and all other variables as predictors. Discrimination is illustrated with a biplot, and using the additional plot arrows from Exercise set 4.

```{r}
# fit linear discriminant analysis
lda.fit <- lda(crime ~ ., data = train)

# the function for lda biplot arrows
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "red", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}

# target classes as numeric
classes <- as.numeric(train$crime)

# plot the lda results
plot(lda.fit, dimen = 2, col = classes, pch = classes)
lda.arrows(lda.fit, myscale = 2)
```

## Predicting crime classification on test set

Using the LDA fit from the train set, we will predict crime category on the test set. We first save the correct classification from the test set in a separate variable, remove the original variable from the test set data frame, and then find predictions based on LDA fit. Finally, we cross-tabulate predictions against correct categories to see classification accuracy.

```{r}
# save correct crime classification in separate variable
correct_classes <- test$crime

# remove crime variable from test data frame
test <- dplyr::select(test, -crime)

# predict classes with test data
lda.pred <- predict(lda.fit, newdata = test)

# cross tabulate the results
table(correct = correct_classes, predicted = lda.pred$class)

```

From the cross-tabulation we can see that the predictions are fairly accurate. Errors are mostly in "adjacent" categories. If anything, classification based on LDA assigns too many observations in the med_low category.

## K-means clustering

For clustering, we will reload the original data, standardize it again, calculate distances between observations, and print a summary of the distance matrix.

```{r}
# reload the data, not strictly necessary as there were no changes to this data frame previously
data("Boston")

# standardizing the data set and changing to data frame
boston_scaled <- as.data.frame(scale(Boston))

# euclidean distance matrix
dist_eu <- dist(boston_scaled)

# print summary of distance matrix
summary(dist_eu)
```
Next, to see examine the optimal number of clusters, we will run the k-means algorithm on the data set with 1 to 10 clusters, calculate and plot total within sum of squares for the different runs.

```{r}
# setting seed for replicability
set.seed(123)

# setting a large number of clusters to explore which number might be optimal
k_max <- 10

# calculate the total within sum of squares
twcss <- sapply(1:k_max, function(k){kmeans(boston_scaled, k)$tot.withinss})

# visualize the results
qplot(x = 1:k_max, y = twcss, geom = 'line')
```

From the drop in the line, two clusters would seem ideal. We will next run k-means clustering with two clusters and visualize the clusters.

```{r}
# k-means clustering
km <- kmeans(boston_scaled, centers = 2)

# set as factor for ggpairs
km$cluster <- factor(km$cluster)

# plot the Boston dataset with clusters
ggpairs(boston_scaled, progress = FALSE, upper = list(continuous = "points"), axisLabels = "none", ggplot2::aes(colour=km$cluster))
```

The scatterplot matrix is again somewhat difficult to read in detail, but some general outlines can be gleaned from the graph. Some clear clustering can be seen in the variables indus (proportion of non-retail business acres per town), nox (nitrogen oxides concentration), rad (index of accessibility to radial highways), and tax (full-value property-tax rate per $10,000). Collectively, this clustering could be indicative of high infrastructure, high industry--and in conjunction--high air pollution areas versus others.

## LDA with clusters as target classes

Next, we will fit and LDA model to the scaled Boston data using clusters from k-means clustering as target classes. We will first run the k-means algorithm (this time for 3 clusters), then include the cluster labels in our scaled data set, and then fit LDA with clusters as the target classes and all other variables as predictors. Finally, we will visualize the results with a biplot.

```{r}
# reload the data, not strictly necessary as there were no changes to this data frame previously
data("Boston")

# setting seed for replicability
set.seed(1)

# standardizing the data set and changing to data frame
boston_scaled <- as.data.frame(scale(Boston))

# k-means clustering with 3 clusters
km <- kmeans(boston_scaled, centers = 3)

# include clusters in the boston data frame
boston_scaled$cluster <- km$cluster

# fit linear discriminant analysis
lda.fit <- lda(cluster ~ ., data = boston_scaled)

# the function for lda biplot arrows
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "red", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}

# target classes as numeric
classes <- as.numeric(boston_scaled$cluster)

# plot the lda results
plot(lda.fit, dimen = 2, col = classes, pch = classes)
lda.arrows(lda.fit, myscale = 3)
```

The three clusters are quite well separated in the biplot. Looking at the arrows for variables in the biplot, we can see that to discriminate cluster one from the other two, variables nox, tax, rad and indus seem most significant. This is fairly similar to what we could tell from the scatterplot matrix for the two clusters in the previous section. We can also see that the variable age (proportion of owner-occupied units built prior to 1940) seems to be most significant in separating cluster two from the others, and that variables zn (proportion of residential land zoned for lots over 25,000 sq.ft), medv (median value of owner-occupied homes in $1000s) and rm (average number of rooms per dwelling) seem to be most significant in separating cluster three from the other two. Cluster number three seems to be connected to size of accommodation, while cluster number two seems to relate to age of housing.

```{r, include=FALSE}
# adding a little timestamp at the end
library(tidyverse)
library(lubridate)
```

```{r, echo=FALSE}
#date()
Sys.time() %>% format("Last updated %a %b %d %H:%M:%S %Y")
```